\documentclass[a4paper,10pt]{letter}
\begin{document}
\pagestyle{empty}
Dear Editors,

We hereby submit our paper \textit{Prior Electrocardiograms not Useful for Predicting Major Adverse Cardiac Events with Machine Learning} for your consideration of publication in Artificial Intelligence in Medicine.

In the pursuit of faster and more efficient assessment of patients with chest pain at the emergency department, AI-based decision support tools are bound to play an increasingly important role. In our paper, we sought to investigate the utility of prior electrocardiograms for improving such algorithms for the purpose of predicting major adverse cardiac events --- one of the leading causes of death in Sweden and worldwide. Prior electrocardiograms are commonly consulted in clinical practice, but our results suggest that they are not useful for machine learning models. This is important for at least three reasons:

\begin{enumerate}
  \item It means that machine learning based decision support tools can maintain predictive accuracy without access to prior electrocardiograms, making them more widely applicable in practice.
  \item It narrows the scope for future efforts to improve such algorithms.
  \item It's an indication that more research is needed to clarify the role of prior electrocardiograms for human analysis, which could potentially be simplified.
\end{enumerate}

Our paper has not been previously published, either in whole or in part, and no similar paper is in press or under review elsewhere. We have no conflicts of interest to declare, and to the best of our knowledge, there are no closely related papers already published.

With best regards,

Axel Nyström

Lund University, Sweden
\end{document}
